---
title: "Development of the FARS (Fatality Analysis Reporting System) Package"
author: "Tony Gojanovic (Coursera course Building R Packages, March 2018)"
date: "April 2, 2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview of the FARS package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This overivew provides a description of an R package called FARS which represents routines for extraction and basic summarization of National Highway Traffic Safety (NHTS) Administration's Fatality Analysis Reporting System data set (which can be found at https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars).

This project was done in partial fullfillment of the Coursera Course "Building R Packages" as developed by John Hopkins University.  The author took this course in March of 2018.

```{r, echo=FALSE, message=FALSE}
library(FARS)
library(dplyr)
library(maps)
```

## FARS Package Content

The FARS package is composed of the following elements or data handling routines:

1. fars_read..............Reading in a data set
2. make _file_name........Creating a filename based on a year input.
3. fars_read_years........Reading in a range of years and creating an aggregate data frame.
4. fars_summarize_years...Simple summary function of accidents by year and month.
5. fars_map_state.........Mapping function of accidents by geographic (state) region.

## Required Packages

The following packages are loaded and required by the package.

```{r Package Loading}
library(FARS)
library(maps)
```

## Data Files

Data files can be found at 

```{r}
list.files(system.file("extdata", package = "FARS"))
```

## fars_read

This function reads in data set from the NHTSA Fatality Analysis Reporting System (FARS).

```{r fars_read function, eval=FALSE}
df<-fars_read("accident_2014.csv.bz2")
```

## make_file_name

This function creates a data file name based on the selected year.

```{r make_filename function, eval=FALSE}
df<-make_filename(2014)
```

## far_read_years

Read the desired yearly data file and create a new data set with two variables, month and year.

```{r far_read_years function, eval=FALSE}
df<-fars_read_years(2013:2015)
```

## fars_summarize_years

This function creates a simple tabular summary of accident counts by year and month.

```{r fars_summarize_years function, eval=FALSE}
fars_summarize_years(2013:2015)
```

## fars_map_state

This function provides a geographic summary of accident locations with a simple state map.

```{r fars_map_state function, eval=FALSE}
fars_map_state(13,2014)
```

Note: For a listing of state numbers, refer to https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/812449

## Limitations

The current package FARS is based on extraction of data elements covering the years 2013-205 e.g. those files that are found within the working directory.  The package could be applied to additional data sets as long as they are of a similar format and can be found in the working directory.

## References

National Highway Traffic Safety Administration's Fatality Analysis Reporting System data set. https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars

